Avoiding the Global Sort: \break A Faster Contour Tree Algorithm

### 2017-05-21 paint-contour-tree.tex
typo corrected

\author{
  Benjamin Raichel\thanks{%
      Department of Computer Science; %
      University of Texas at Dallas; %
      Richardson, TX, 75080, USA; %
      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}}.
  The work was done when this author was an intern
  at Sandia National Laboratories, Livermore.}
  \and
  C. Seshadhri\thanks{%
        Department of Computer Science; %
      University of California, Santa Cruz; %
      Santa Cruz, CA, 95064, USA; %
      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
  }
}

@@

Consider $f: \MM \to \RR$, where $\MM$ is the triangulation of a $d$-dimensional ball with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$, see \Def{deg}. 
The degree assumption can be achieved via vertex unfolding, as discussed later,
%We define this formally later, but it suffices to think of this as a general position assumption.
and is standard in this literature~\cite{kobps-ctsssit-97}.)
% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

% \Sesh{Almost everything below is new/reordered. Please read through.}
All previous algorithms begin by sorting (at least) the critical points. Can we beat this sorting bound for certain instances,
and can we characterize which inputs are hard? Intuitively, points that are incomparable in the contour tree
do not need to be compared. Look at \Fig{relative} to see such an example. All previous algorithms waste time
sorting all the maxima. Also consider the surface of \Fig{sorting}. The final contour tree is basically two binary
trees joined at their roots, and we do not need the entire sorted order of critical points
to construct the contour tree.

% \begin{figure}[h]\centering
%     \includegraphics[width=.26\linewidth]{figs/spikey1}%
%     \hspace{.35in}
%     \includegraphics[width=.26\linewidth]{figs/spikey2}%
%     \hspace{.4in}
%     \includegraphics[width=.215\linewidth]{figs/spikeyTree}
%     \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
%     \label{fig:relative}
% \end{figure}

@@

% First, we require some technical definitions.

% In a contour tree $T$, a \emph{leaf path} is simply a path in $T$
% containing a leaf, which is also monotonic in the height values of its vertices.

@@

\begin{theorem} 
\label{thm:main-alg} 
%Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies that all critical points have distinct values. 
%Our algorithm does not need this, but it simplifies the proof and presentation.} 
(Consider the input setting of \Thm{main-corr}.)
There is a deterministic algorithm to compute the contour tree, $T$, whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
In particular, comparisons are only made between pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
\end{theorem}

@@

The first formal result was an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalyi \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-03} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
% Carr's thesis shows relationships between monotone paths
% in $\MM$ and its contour tree $\cC(\MM)$.

@@

% This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
% has a well-defined height, including the infinite facet, which is required to be a boundary facet.  However, 
% we allow for other boundary facets, to capture the resulting surface pieces after our algorithm makes a horizontal cut.

@@

% We categorize points depending on the local changes in topology.
% 
% \begin{definition} \label{def:points} A point $x$ is categorized as follows.
% \begin{asparaitem}
% 	\item \emph{Regular:} Both Morse up-degree and down-degrees are $1$.
% 	\item \emph{Maximum:} Morse up-degree is $0$.
% 	\item \emph{Minimum:} Morse down-degree is $0$.
% 	\item \emph{Morse Join:} Morse up-degree is strictly greater than $1$.
% 	\item \emph{Morse Split:} Morse down-degree is strictly greater than $1$.
% \end{asparaitem}
% Non-regular points are \emph{critical}. Morse joins and splits are \emph{saddles}.
% Maxima and minima are \emph{extrema}.
% \end{definition}
%

@@

This equivalence relation gives a set of \emph{contour classes}. 
%By continuity, a regular point is present on a single contour,
%and we can partition the regular contours into these classes. 
Every such class maps to intervals
of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
to be created at $x_i$ and destroyed at $x_j$.

@@

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but alternative definitions of the contour tree
in~\cite{csa-cctad-03} imply this is a tree. 
Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
%

@@

%\pagebreak
\section{A tour of the new contour tree algorithm} \label{sec:approach}
We provide a high-level description of the main ideas of subsequent sections.

@@

\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
(When formally defined in \Sec{rain}, extremum dominant complexes will allow additional trivial minima or maxima which are a small complication resulting from the following cutting procedure.)
We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
the wet part, and recurse on the dry parts. This is carefully done to ensure that the total complexity
of the pieces is linear (see \Sec{rain} for details).
% 
% 
% Each additional cut increases the complexity of the surface,
% so care is required to keep the total complexity linear. This is achieved by alternately ``raining"
% upwards and downwards.  We finally get a collection of extremum dominant pieces.
% We prove a simple contour surgery theorem that builds the contour tree of $\MM$
% from the contour trees of the various pieces created.

@@

% For a fast running time, we need to find ripe vertices
% quickly. Thus, the formal algorithm requires a 
% number of data structures, described below.

@@

\begin{theorem} \label{thm:jordan} $\cut(\MM,\phi)$ consists of two
disconnected simplicial complexes, each of which
is the triangulation of $d$-dimensional ball.
\end{theorem}

@@

% 
% \begin{lemma} \label{lem:blahblah}
%  Testing 123 \RefProofInAppendix{blahblah}
% \end{lemma}
% 
% \begin{proof:in:appendix:e}{\ref{lem:blahblah}}{blahblah}
%  blah balh
% \end{proof:in:appendix:e}

@@

Since $T_r$ is not used up, it must mean that all $p$'s are full, as $r$ is the root and so $T_r$ tokens can be placed on any path which is not full. 
Consider any $p$, and let $K_p$ be the set of tokens placed at $p$. Each token is labeled $\log |H_v|$,
for some $v \in \loss(\hat{p})$, where $\hat{p}$ is an ancestor of $p$ or $p$ itself (as descendants do not place tokens on ancestors). By \Cor{geometric},
for every $w \in p$, $\log |H_w| = O(\log |H_v|)$. Because $p$ is full,
there are exactly $|p|$ tokens on $p$. Thus, $\sum_{w \in p} \log |H_w|$
is $O(V_p)$, where $V_p$ is the sum of token values in $K_p$. 
Thus summing over all $p$ in $\buff$ completes the proof. 
Specifically, $\sum_{p\in \buff} \sum_{w \in p} \log |H_w| = O(\sum_{p\in \buff} V_p) = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$

% \textcolor{red}{\huge Proof missing. Potentially useful observations below}
%  
%  Let $r$ be the root of $\buff$. First observe that $\loss(r) \geq |r|/100$.  Suppose otherwise, then in order for the first condition of \Def{buff} to be satisfied for $\buff$, the rooted subtree of some child of $r$, call it $\buff'$, must satisfy $\sum_{p\in \buff'} \loss(p) \geq \frac{1}{100} \sum_{p\in \buff'} |p|$.  However, this would contradict the second condition of \Def{buff}.
%  
%  Consider any node $q\in \buff$.  Then for any vertex $v\in \loss(q)$, we have $|H_v| = \Omega(|H_q|)$, since $q$ is unsupported and hence by definition less than $|H_q|/100$ vertices of $H_q$ appear in $\loss(q)$.
%  
%  \begin{align*}
%   \sum_{p\in \buff}\sum_{v\in p} \log |H_v| &=  O\left( \sum_{p\in \buff}\sum_{v\in p} \log |H_p| \right)   &\mbox{definition of short} \\
%   &= O\left( \sum_{p\in \buff} |p| \log |H_p| \right)   &\mbox{} \\
%   &= ????   &\mbox{buffered is not sufficient to get to next line} \\
%   &= O\left( \sum_{p\in \buff} |\loss(p)| \log |H_p| \right)   &\mbox{} \\
%   &= O\left( \sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| \right)
%  \end{align*}
%  
%  \Cor{geometric} should be used somewhere.
\end{proof}

@@

% 
% \begin{definition}\label{def:pruningAShrub}
%  Let $\cU$ be a shrub in $\cF$. Consider a reverse topological ordering (i.e.\ leaves first) of the nodes in $\cU$, and use it to define a total ordering.  
%  We compute a pruning of $\cU$ (see \Def{prune}), by iteratively removing all buffered sub-shrubs (see \Def{buff}), where buffered sub-shrubs whose roots are earlier in the total ordering are pruned first. We use $\pruned$ to denote the resulting sub-shrub, which we refer to as the \emph{buffer-pruned} sub-shrub of $\cU$.
% \end{definition}
% 
The goal of this section is to prove the following lemma.
% , which ultimately allows us to pay for pruned shrubs.

@@

The proof of \Lem{cu-root} just combines all these claims, and essentially
repeats the argument in the previous claim. We apply \Clm{exp-A} and \Clm{charge}
to every child of the root $r$. Denoting the children as $q_1, q_2, \ldots$,
we get that $\sum_{p \in \pruned_{q_i}} |p| \leq A_{q_i}/20$.
Summing this over the children, $\sum_{p \in \pruned, p \neq r} |p| \leq A_r/20$.
By \Clm{weights}, $|H_r| \geq 17A_r/20 - 3|r|$. Since $r$ represented a short path, 
we have the bound $|r| \leq \sqrt{|H_r|}/100$. Thus, $|H_r| = \Omega(A_r)$.
We combine these bounds to show that $\sum_{p \in \pruned} |p| = O(|H_r|)$.
% 
% 
% 
% \begin{proof}[Proof of \Lem{cu-root}] We first pro 
% We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
% Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
% then by \Cor{chargeCor}, $\sum_{p \in \cU} |p| \leq W_{r}/20+|r|$.
% By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$, and so combining these inequalities gives,
% \[
% % \sum_{p \in \cU} |p| \leq \pth{|\redH_r|+2\sum_{p \in \cU} |p|}/20+|r| \Rightarrow 
%  \sum_{p \in \cU} |p|\leq \frac{10}{9} \pth{|\redH_r|/20+|r|} \leq 
%  \frac{10}{9}\pth{|H_r|/20+\sqrt{|H_r|}/100} \leq |H_r|/10.
% \]
% 
% We now prove that for any $q$ parent of $q'$ (other than $r$), $W_q \geq (4/3)W_{q'}$.
% Suppose not. Let $p, p'$ be the counterexample furthest from the root,
% where $p$ is the parent of $p'$. 
% Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
% We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
% Combining this with \Clm{loss} gives, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
% Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
% Hence, $p, p'$ is not a counterexample, and more generally, there is no counterexample.
% That completes the whole proof.
% \end{proof}
% 
% 
% \textcolor{red}{\huge The rest of this subsection needs updating}
% 
% We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
% $q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
% By definition, $|\redH_p| = \sum_{q \in \cU_p} R_q$. 
% Note that $R_p$ can be negative. 
% Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_q$, 
% where $\cU_p$ is the subtree of $\cU$ rooted at $p$.
% Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.
% 
% 
% \begin{claim} \label{clm:loss} For any path $p\in \cU$, $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$.
% \end{claim}
% 
% \begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} (R^+_q - R_q)$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
% Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
% Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
% Traverse $\pmax(T)$ from $q'_i$ to $q$.
% Other than $q$, all other nodes encountered are in the tall support chain of $q'_i$ and hence
% do not affect its reduced heap.
% The vertices of $\redH_{q'_i}$ that are deleted
% are exactly those present in the path $q$. 
% Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
% theses reduced heaps do not have an ancestor-descendant relationship.
% Therefore when $R_q$ is negative, it is at most by $2|q|$.
% We sum over all $q$ to complete the proof.
% \end{proof}
% 
% \ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
% $D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
% to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.
% 
% 
%  and traverse $\pmax(T)$ up to $p$. Suppose we encounter
% the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
% are removed from $H_p$ are exactly the vertices that lie in these paths. 
% We split this contribution into tall and short paths, observing that the latter
% are all unsupported. Hence, we can (crudely) upper bound the number of vertices
% removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
% Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
% }
% 
% 
% The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
% We stress that the all the previous work is mostly the setup for this claim.
% 
% 
% \begin{corollary} \label{cor:chargeCor}
%  Let $r$ be the root of $\cU$, and suppose that for any paths $q,q' \in \cU\setminus \{r\}$, where $q$ is a parent of $q'$ in $\cU$, $W_q \geq (4/3)W_{q'}$.
%  Then $\sum_{p \in \cU} |p| \leq W_r/20+|r|$.
% \end{corollary}
% \begin{proof}
%  Let $c_1, \dots, c_m$ be the children of $r$ in $\cU$.  By definition, $W_r = \sum_i W_{c_i}+R^+_r \geq \sum_i W_{c_i}$.
%  By \Clm{charge}, for each $c_i$ we have $W_{c_i}/20\geq \sum_{p \in \cU_{c_i}} |p|$.  Combining these to facts yields the claim.
% \end{proof}
% 
%

% \begin{claim} \label{clm:charge} Fix $p \in \cU$. Suppose for any $q,q' \in \cU_p$
% where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
% \end{claim}
% 
% \begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
% We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
% by a charge redistribution scheme. Assume that each $q \in \cU_p$
% starts with $\sqrt{W_p}/99$ units of charge. We redistribute this charge over all
% nodes in $\cU_p$, and then calculate the total charge. For $q \in \cU_p$, spread
% its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
% give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
% 
% After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p$ 
% $q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
% equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
% as $a_0$ is an unsupported short path and hence $W_{a_0}\geq 1$. 
% $$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
% The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
% \end{proof}

% We wrap it all up by proving part (ii) of \Lem{cu-root}.
%
% \begin{claim} $\sum_{p \in \cU} |p| \leq |H_{r(\cU)}|/10$.
% \end{claim}
% 
% % \begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
% % Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
% % By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.
% % 
% % We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
% % Suppose not. Let $p, p'$ be the counterexample furthest from the root,
% % where $p$ is the parent of $p'$. 
% % Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
% % We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
% % By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
% % Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
% % Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
% % That completes the whole proof.
% % \end{proof}

@@

% \begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
% \end{claim}
% 
% \begin{proof} 
% Pick $p \in \cL'$. 
% As we traverse the tall support chain of $p$, there
% are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
% a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
% is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
% $|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
% Since all the vertices lie in tall paths,
% we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
% Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
% %and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
% 
% Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
% $v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e.\ it contains at most one path other than $p$).
% First observe that any two paths in $S$ must be unrelated (i.e.\ $S$ is an anti-chain), 
% since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
% However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
% paths which share descendant leaves in $T$ (i.e.\ their heaps are tracking the same color) must have an ancestor-descendant relationship.
% 
% In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
% Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
% \end{proof}

% The first term (unsupported short paths) can be charged to the second term (supported short paths).
% This is where the critical \Lem{cu-root} plays a role.

@@

At this point we can now see what the path decomposition referenced in theorem statement should be.  It is just the union of all the maximum 
path decompositions across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\cJ(\MM_i))$.
Since all procedures besides computing the join trees take linear time in the size of the input complex, we can therefore 
compute the contour tree in time
\[
O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\cJ(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}
=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
\]

%\Thm{rain-time} proves that $\rain$ takes $O(|\MM|)$ time

%\Lem{rain-1} proves that output of $\rain$ are extemum dominant.

%\Clm{rain-reeb} shows that given the contour trees of the output manifolds of $\rain$, the full contour tree can be constructed in $O(|\MM|)$ time:

%\Thm{contour-tree} implies that given an $\jc(\MM)$ for an etremum dominant $\MM$, we can obtain $\cC(\MM)$ in by simply sticking on the non-dominant minima at their respective splits (which can easily be done in linear time). 
\end{proof}

@@

% \textcolor{red}{
% In the example below, I assumed that the maxima are already considered ``processed'', so that there are not additional trivial steps.  
% This actually makes more sense since the $att()$ should really only ever point to already processed vertex.  Therefore, I suggest 
% changing line 3 of init to ``Mark all maxima as processed.  Mark all other critical $v$ as unprocessed.''
% Perhaps line 4 needs to be changed also?
% }

>>>>

\author{
  Benjamin Raichel\thanks{%
      Department of Computer Science; %
      University of Texas at Dallas; %
      Richardson, TX, 75080, USA; %
      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}}.
}
  \and
  C. Seshadhri\thanks{%
        Department of Computer Science; %
      University of California, Santa Cruz; %
      Santa Cruz, CA, 95064, USA; %
      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
  }
}

@@

Consider $f: \MM \to \RR$, where $\MM$ is the triangulation of a $d$-dimensional ball with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$, see \Def{deg}. 
The degree assumption can be achieved via vertex unfolding, as discussed later,
and is standard in this literature~\cite{kobps-ctsssit-97}.)
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

All previous algorithms begin by sorting (at least) the critical points. Can we beat this sorting bound for certain instances,
and can we characterize which inputs are hard? Intuitively, points that are incomparable in the contour tree
do not need to be compared. Look at \Fig{relative} to see such an example. All previous algorithms waste time
sorting all the maxima. Also consider the surface of \Fig{sorting}. The final contour tree is basically two binary
trees joined at their roots, and we do not need the entire sorted order of critical points
to construct the contour tree.

@@

@@

\begin{theorem} 
\label{thm:main-alg} 
(Consider the input setting of \Thm{main-corr}.)
There is a deterministic algorithm to compute the contour tree, $T$, whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
In particular, comparisons are only made between pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
\end{theorem}

@@

The first formal result was an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalyi \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-03} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.

@@

@@

@@

This equivalence relation gives a set of \emph{contour classes}. 
Every such class maps to intervals
of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
to be created at $x_i$ and destroyed at $x_j$.

@@

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but alternative definitions of the contour tree
in~\cite{csa-cctad-03} imply this is a tree. 
Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.

@@

\section{A tour of the new contour tree algorithm} \label{sec:approach}
We provide a high-level description of the main ideas of subsequent sections.

@@

\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
(When formally defined in \Sec{rain}, extremum dominant complexes will allow additional trivial minima or maxima which are a small complication resulting from the following cutting procedure.)
We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
the wet part, and recurse on the dry parts. This is carefully done to ensure that the total complexity
of the pieces is linear (see \Sec{rain} for details).

@@

@@

\begin{theorem} \label{thm:jordan} $\cut(\MM,\phi)$ consists of two
disconnected simplicial complexes, each of which
is the triangulation of a $d$-dimensional ball.
\end{theorem}

@@

@@

Since $T_r$ is not used up, it must mean that all $p$'s are full, as $r$ is the root and so $T_r$ tokens can be placed on any path which is not full. 
Consider any $p$, and let $K_p$ be the set of tokens placed at $p$. Each token is labeled $\log |H_v|$,
for some $v \in \loss(\hat{p})$, where $\hat{p}$ is an ancestor of $p$ or $p$ itself (as descendants do not place tokens on ancestors). By \Cor{geometric},
for every $w \in p$, $\log |H_w| = O(\log |H_v|)$. Because $p$ is full,
there are exactly $|p|$ tokens on $p$. Thus, $\sum_{w \in p} \log |H_w|$
is $O(V_p)$, where $V_p$ is the sum of token values in $K_p$. 
Thus summing over all $p$ in $\buff$ completes the proof. 
Specifically, $\sum_{p\in \buff} \sum_{w \in p} \log |H_w| = O(\sum_{p\in \buff} V_p) = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
\end{proof}

@@

% 
% \begin{definition}\label{def:pruningAShrub}
%  Let $\cU$ be a shrub in $\cF$. Consider a reverse topological ordering (i.e.\ leaves first) of the nodes in $\cU$, and use it to define a total ordering.  
%  We compute a pruning of $\cU$ (see \Def{prune}), by iteratively removing all buffered sub-shrubs (see \Def{buff}), where buffered sub-shrubs whose roots are earlier in the total ordering are pruned first. We use $\pruned$ to denote the resulting sub-shrub, which we refer to as the \emph{buffer-pruned} sub-shrub of $\cU$.
% \end{definition}
% 
The goal of this section is to prove the following lemma.

@@

The proof of \Lem{cu-root} just combines all these claims, and essentially
repeats the argument in the previous claim. We apply \Clm{exp-A} and \Clm{charge}
to every child of the root $r$. Denoting the children as $q_1, q_2, \ldots$,
we get that $\sum_{p \in \pruned_{q_i}} |p| \leq A_{q_i}/20$.
Summing this over the children, $\sum_{p \in \pruned, p \neq r} |p| \leq A_r/20$.
By \Clm{weights}, $|H_r| \geq 17A_r/20 - 3|r|$. Since $r$ represented a short path, 
we have the bound $|r| \leq \sqrt{|H_r|}/100$. Thus, $|H_r| = \Omega(A_r)$.
We combine these bounds to show that $\sum_{p \in \pruned} |p| = O(|H_r|)$.

@@

@@

At this point we can now see what the path decomposition referenced in theorem statement should be.  It is just the union of all the maximum 
path decompositions across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\cJ(\MM_i))$.
Since all procedures besides computing the join trees take linear time in the size of the input complex, we can therefore 
compute the contour tree in time
\[
O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\cJ(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}
=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
\]
\end{proof}

@@
